---
title: "Adult Census Income"
author: "Harsh Navin Gupta"
date: "10 January 2020"
output: pdf_document
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dataset Analysis
The **Adult Census Income Dataset**, contains features that are being used to predict whether a person, earns >50K or <=50K.  
The dataset, contains an observation for every individual person, and the feature to be predicted is the *income*, which is composed of two values : *<=50K, >50K*.  
The **Adult Census Income Dataset** contains the following features :  
1. **Age** : Stores the age of the individual.  
2. **Workclass** : Stores the type of employment of the individual, whether he/she is a federal emplyee, private employee, or has his/her own buisness.  
3. **fnlwgt** : Stores the sampling weight.  
4. **Education** : Stores the highest degree of education, held by the individual.  
5. **Education-Num** : Stores the number of years of education completed by the individual.  
6. **Marital-Status** : Stores the marital status of the individual, whether they are married, divorced,etc.  
7. **Occupation** : Stores a short descriptor about the type of job of the individual.  
8. **Relationship** : Stores the relationship which the individual holds, if he/she is a part of a family.  
9. **Race** : Stores the race of the individual.  
10. **Sex** : Stores the sex of an individual.  
11. **Capital-Gain**  
12. **Capital-Loss**  
13. **Hours-Per-Week** : Stores the number of hours the individual works in a week.  
14. **Native-Country** : Stores the country to which the individual natively belongs.  

## Dataset Download
The dataset consists of two files, that have to be downloaded.  
1. **adult.data** : This is a CSV file, that contains the training data.  
2. **adult.test** : This is a CSV file, that contains the testing data.  

The link to download the **adult.data** file is :  
<https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data>    

The link to download the **adult.test** file is :  
<https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.test>    

First, we check for the library that are required, and load the required libraries.    

```{r lib_check, warning=FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", repos = "http://cran.us.r-project.org")
```  

Now, we create a vector of column names, for the data frames to be used.  

```{r col_names}
col_names <- c("age","workclass","fnlwgt","education",
               "education_num","marriage","occupation",
               "relationship","race","sex","capital-gain",
               "capital-loss","hours",
               "country","income")
```  

Now, we first download the file **adult.data** and create our *training set*.  

```{r train_down}
#Link To Download adult.data
train_link <- "https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data" 
train_set <- read_csv(train_link,col_names = FALSE)
train_set <- setNames(train_set,col_names)
head(train_set)
```  

Now, we proceed to download the file **adult.test** and create our *testing set*.  
```{r test_down}
#Link To Download adult.test
test_link <- "https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.test"
test_set <- read_csv(test_link,col_names = FALSE,skip = 1)
test_set <- setNames(test_set,col_names)
head(test_set)
```  

## Data Preprocessing  
The columns of the dataset, *workclass, occupation, and native-country*, contain unknown values.  

First, we discard all observations with unknown values *(represented by ?)*, from the training set.  

```{r train_unknown}
train_set <- train_set %>% 
  filter(workclass != "?") %>%
  filter(occupation != "?") %>%
  filter(country != "?")
```  

Next, we discard all observations with unknown values *(represented by ?)*, from the testing set.  

```{r test_unknown}
test_set <- test_set %>% 
  filter(workclass != "?") %>%
  filter(occupation != "?") %>%
  filter(country != "?")
```  

First, we calculate the *Average Weekly Working Hours* of all the individuals.  

```{r mean_work}
mean_hours <- mean(train_set$hours)
```  

Now create a new column named *work_hour_group*, which can have anyone of two values :  
1. **Lower** : If the works hours per week of the individual are lower than the average weekly work hours.  
2. **Higher** : If the works hours per week of the individual are higher than the average weekly work hours.  

We initially make the changes to *Training Set*.  

```{r train_work}
train_set <- train_set %>%
  mutate(work_hour_group = ifelse(hours < mean_hours,"Lower","Higher"))
```  

Then, we make the changes to the *Testing Set*.  

```{r test_work}
test_set <- test_set %>%
  mutate(work_hour_group = ifelse(hours < mean_hours,"Lower","Higher"))
```  

Next, we categorise the countries in the dataset, as *Developed Countries (D)* or *Under Development Countries*.  
The following countries are categorised as *Developed Countries* :  
* Germany  
* England  
* France  
* Japan  
* Canada  
* United-States  
* Ireland  
* Italy  
The remaining countries are classified as *Under Development Countries (UD)*.  

```{r dev_c}
#Categorising Countries As Developed & Under Development
dc <- c("Germany","England","France","Japan",
        "Canada","United-States","Ireland","Italy")
```  

Initially we make changes to the *Training Set*.  

```{r train_cntry}
train_set <- train_set %>% 
  mutate(country_type = ifelse(country %in% dc,"D","UD"))
```  

Then, we make changes to the *Testing Set*.  

```{r test_cntry}
test_set <- test_set %>% 
  mutate(country_type = ifelse(country %in% dc,"D","UD"))
```  

For the purpose of ease of training ML Models on the training set, we add a new column *y* to both training and testing sets, which contains the following values :  
1. **0** : If the individual income is equal to <=50K  
2. **1** : If the individual income is equal to >50K  

Initially, we make changes to the *Training Set*.  

```{r train_y}
#Creating New Column y
train_set <- train_set %>% mutate(y = ifelse(income == "<=50K",0,1))
train_set <- train_set %>% mutate(y = factor(y))
```  

Then we make changes to the *Testing Set*.  

```{r test_y}
#Creating New Column y
test_set <- test_set %>% mutate(y = ifelse(income == "<=50K.",0,1))
test_set <- test_set %>% mutate(y = factor(y))
```  

## Exploratory Data Analysis  

```{r total_ppl}
#Total People In The Training Set
nrow(train_set)
```  

```{r total_ctry}
#Total Countries In The Training Set
n_distinct(train_set$country)
```  

```{r total_sex}
#Total Male & Females In The Training Set
train_set %>% group_by(sex) %>% 
  summarise(count = n()) %>% knitr::kable()
```  

### Income Over Sex  
Here, we visualise the distribution of the income, over Males & Females, which are present in the training set.

```{r in_sx}
train_set %>% group_by(sex,income) %>% 
  summarise(count = n()) %>% knitr::kable()
```

```{r in_sx_two}
train_set %>% group_by(sex,income) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(income,count,fill = sex)) + 
  geom_bar(stat = "identity",position = position_dodge2()) + 
  xlab("Sex") + ylab("Count") + 
  theme_gdocs() + ggtitle("Income Over Sex")
```  

Here, it is observed that, the proportion of people earning <=50K is greater than people earning >50K for both the genders.  
It can also be clearly observed that total number of Males are greater than the total number of Females.  

### Income Over Different Races
Here, we analyse the distribution of the income, among individuals belonging to different races.  

```{r in_race}
train_set %>% group_by(race,income) %>% 
  summarise(count = n()) %>% knitr::kable()
```  

```{r in_race_two}
train_set %>% group_by(race,income) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(income,count,fill = race)) + 
  geom_bar(stat = "identity",position = position_dodge2()) + 
  theme_gdocs() + 
  xlab("Income") + ylab("Count") + 
  ggtitle("Income Over Different Races")
```  

Here, it can clearly observed that the dataset is dominated by people belonging to the *White* race.  
But, it can be observed in general, that the people with income <=50K are greater in number than people with income >50K, across all the races.  

### Income Over Education  
Here we analyze the distribution of the income, in comparison to the highest degree of education held by an individual.  

```{r in_ed}
train_set %>% group_by(education,income) %>% 
  summarise(count = n()) %>% knitr::kable()
```  

```{r in_ed_two}
train_set %>% group_by(education,income) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(reorder(education,count),count,fill = income)) + 
  geom_bar(stat = "identity",position = position_dodge2()) + 
  coord_flip() + 
  theme_gdocs() + 
  xlab("Education") + ylab("Count") + 
  ggtitle("Income Over Education")
```  

Here it can be clearly observed, that the people who hold Master's degree, or a Doctrate, or have attended Prof-School, have higher proportion of them, earning >50K, in comparison to other degrees, where the proportion of people earning, <=50K is greater.  

It can also be observed that most people, in the dataset, are *High School Graduates**.  

### Income Over Different Occupations  
Here we analyse the distribution of incomes of the individuals in the database, on the basis of their occupation.  

```{r in_occ}
train_set %>% group_by(occupation,income) %>% 
  summarise(count = n()) %>% knitr::kable()
```  

```{r in_occ_two}
train_set %>% group_by(occupation,income) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(reorder(occupation,count),count,fill = income)) +
  geom_bar(stat = "identity",position = position_dodge2()) + 
  coord_flip() + 
  ggtitle("Income Over Different Occupations") + 
  xlab("Occupation") + ylab("Count") + theme_gdocs()
```  

Here, in general, it is observed the people earning <=50K are greater than the number of people earning >50K.  
However, it can also be seen that occupations such as *Prof-speciality* and *Exec-managerial*, usually requiring a *Master's Degree/ Doctrate*, have a comparitively little difference in the number of people earning <=50K & >50K.  
This observation, also acts as a supporter to the conclusion drawn using the Income Over Education visualisation.  

### Income Over Hours Per Week  
In this section, we analyse the income distribution, for the people who work less than the average hours per week, and more than the average hours per week.  
We also analyse the Hours Per Week, with respect to the Gender of the individual, to determine, whether there is difference in the trend, if considered on a gender basis.  

```{r in_wh}
train_set %>% group_by(work_hour_group,income) %>% 
  summarise(count = n()) %>% knitr::kable()
```  

```{r in_wh_two}
train_set %>% group_by(work_hour_group,income) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(work_hour_group,count,fill = income)) + 
  geom_bar(stat = "identity",position = position_dodge2()) + 
  xlab("Work Hour Group") + 
  ylab("Count") + 
  ggtitle("Income Over Work Hour Group") + 
  theme_gdocs()
```  

It can be clearly observed that for the people who work *Greater than the Average Hours Per Week*, the proportion of people earning <=50K, still remains larger, than the people earning >50K.  
However, it can also be seen that for people working *Greater than the Average Hours Per Week*, the difference in numbers is very small, when compared to the people working *Less Than The Average Hours Per Week*.  

Now, we compare this trend, by also considering, the gender of the individuals.  
```{r in_wh_three}
train_set %>% group_by(work_hour_group,income,sex) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(work_hour_group,count,fill = income)) + 
  geom_bar(stat = "identity",position = position_dodge2()) + 
  xlab("Work Hour Group") + 
  ylab("Count") + 
  ggtitle("Income Over Work Hour Group & Sex") + 
  theme_gdocs() + facet_grid(.~sex)
```  

It can be seen that the *Gender* does not act as a *Bias*, in anyway, and the trend observed earlier, still remains true.  

### Income Over Maritial Status  
Here, we analyse the distribution of income, based on the individual's maritial status.  

```{r in_mr}
train_set %>% group_by(marriage,income) %>% 
  summarise(count = n()) %>% knitr::kable()
```  

```{r in_mr_two}
train_set %>% group_by(marriage,income) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(marriage,count,fill = income)) + 
  geom_bar(stat = "identity",position = position_dodge2()) + 
  xlab("Maritial Status") + 
  ylab("Count") + 
  ggtitle("Income Over Maritial Status") + 
  theme_gdocs() + coord_flip()
```  

In the visualisaton, it can be clearly observed, that the chances of a person earning <=50K of not being married, are higher than that of being married.  
The highest number of people earning <=50K, have never married.  

### Income Over Relationship  
Here, we verify the observations that have been observed in the previous visulation.  

```{r in_r}
train_set %>% group_by(relationship,income) %>% 
  summarise(count = n()) %>% knitr::kable()
```  

```{r in_r_two}
train_set %>% group_by(relationship,income) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(relationship,count,fill = income)) + 
  geom_bar(stat = "identity",position = position_dodge2()) + 
  xlab("Relationship") + 
  ylab("Count") + 
  ggtitle("Income Over Relationship") + 
  theme_gdocs()
```  

It can be clearly observed that the highest number of people earning <=50K, are current not in a family, confirming our previous conclusion, *Person earning <=50K, has a higher probability of being never married*.  

### Income Over Country Type
Here, we analyse whether the distribution of incomes, difference in *Developed vs Under Development Countries*. 

```{r in_cn}
train_set %>% group_by(country_type,income) %>%
  summarise(count = n()) %>% 
  ggplot(aes(country_type,count,fill = income)) +
  geom_bar(stat = "identity",position = position_dodge2()) +
  ggtitle("Income Over Country Type") +
  xlab("Country Type") + 
  ylab("Count") + theme_gdocs()
```  
Here, it can clearly be seen, that most of the individuals of the dataset, are native to a developed country.  
Secondly, be it developed or under development country, the number of people earning <=50K are significantly higher than number of people earning >50K.  

## Model Fitting  
Based on the *Exploratory Data Analysis*, we will use only the following columns for the training of the *ML Models*.  
1. **sex**  
2. **age**  
3. **occupation**  
4. **education**  
5. **relationship**  
6. **race**  
7. **marriage**  
8. **hours**  
9. **country**  
10. **y**  

```{r sel_col}
selected_features <- c("sex","age","occupation",
                       "education","relationship","race",
                       "marriage","hours","country","y")
```  

```{r t_t_col}
#Selecting Columns From Training Set
train_set <- train_set %>% select(selected_features)

#Selecting Columns From Testing Set
test_set <- test_set %>% select(selected_features)
```  

### Using Logistic Regression  
First, we define *K Fold Cross Validation*. Here, we define, that only *10 Times (K = 10)*, will be performed, and the validation set *will be 10% of the training set (p = 0.9)*. 

```{r k_fold}
control <- trainControl(method = "cv",number = 10,p = 0.9)
```  

Now, we fit the *Logistic Regression Model*, to the training set, using the train() function of the caret package.  

```{r lg_fit, message=FALSE, warning=FALSE}
lga_fit <- train(y ~ .,data = train_set,method = "glm",trControl = control)
```  

Now, we predict the income for the *Testing Set* by making of the predict() function.  

```{r lg_pred}
y_hat <- predict(lga_fit,test_set)
```  

We can analyse the performance of the Model, by making use of the Confusion Matrix.  

```{r lg_cm}
confusionMatrix(y_hat,test_set$y)
```  

The *Accuracy* for the *Logistic Regression Model*.  
```{r lg_acc}
confusionMatrix(y_hat,test_set$y)$overall[1]
```  

### Using Classifier Tree  
Here, we make of *tuneGrid* argument, to fit the Classifier tree, for multiple *cp (Complexity Parameter)* values, and fit the model using the *train()* function of the caret package.  

```{r ct_fit}
tree_fit <- train(y ~ .,data = train_set,
                  method="rpart",
                  tuneGrid=data.frame(cp=seq(0,0.05,0.002)))
```  

Now, we plot the *cp (Complexity Parameter) VS Accuracy*.  

```{r ct_plot}
ggplot(tree_fit,highlight = TRUE) + theme_gdocs()
```  

The best value of *cp* is :  

```{r ct_cp}
tree_fit$bestTune
```  

Now, we plot the *Best Fit Classifier Tree Model*.  

```{r ct_plot_best}
plot(tree_fit$finalModel,margin = 0.01)
text(tree_fit$finalModel,cex = 0.6)
```  

Now, we predict the income for the *Testing Set* by making of the predict() function.  

```{r ct_pred}
y_hat <- predict(tree_fit,test_set)
```  

We can analyse the performance of the Model, by making use of the Confusion Matrix.  

```{r ct_cm}
confusionMatrix(y_hat,test_set$y)
```  

The *Accuracy* for the *Classifier Tree Model*.  
```{r ct_acc}
confusionMatrix(y_hat,test_set$y)$overall[1]
```  

## Conclusion  
For the *Adult Census Income* dataset, both the *Logistic Regression & Classifier Tree* models, perform equally good, by making use of *Accuracy* as the comparison parameter, and hence any of the model can be used for predicting the income of an individual.  

## Github Link  
<https://github.com/guptaharshnavin/Adult_Census_Income>  